Supporting Materials for \textit{Category-based and location-based volitional covert attention affect memory at different timescales}

### 2023-03-26 paper/main.tex
finished draft of results section and supplement

\newcommand{\gazeLocations}{S1}
\newcommand{\excludedTrials}{S2}
\newcommand{\ratingsByCategory}{S3}

@@

Our brains' cognitive systems detect and exploit patterns in our prior and
ongoing experiences, enabling us to function and adapt in an ever-changing
world. However we do not attend to or treat all types of remembered or incoming
information equally, and our ability to flexibly adapt our thinking and
behaviors can vary markedly with the specific set of concepts or tasks relevant
to a given setting or situation~\citep{ChunTurk07, AlyTurk17, HardNade09,
RangRitc12}. There is also substantial variability across people with respect
to which aspects of experience (sensory, social, emotional, etc.) are noticed,
discriminated between, and acted upon~\citep{HuntEtal89}. This implies that the
same physical (objective) experience may give rise to very different perceived
(subjective) experiences across people~\citep{FreeSimo11, ChanEtal18}.

The aspects of our experience we attend may be under our volitional control or
may be unconscious or automatic~\citep{JacoEtal92}. Both volitional and
unconscious attention may be expressed overtly, for example through intentional
eye movements~\citep{HoffSubr95} or covertly, without any volitional physical
change~\citep{EngbKlie03}. Prior work has explored the similarities and
differences in the neural basis of overt versus covert
attention~\citep{PosnEtal87, HuntKing03} as well as the behavioral and neural
underpinnings of volitional versus unconscious attention~\citep{DijkAart10} and
their differential effects on memory. There is a general consensus that
sustained volitional attention enhances memory relative to unconscious
attentional processes~\citep{UncaEtal11, GoloEtal13}. However, volitional
attention takes many forms, such as attention to particular spatial locations
or attention to particular visual features or other stimulus properties. How
different \textit{types} of volitional attention combine (or compete) to
enhance memory remains an open question. Volitional covert attention is of
particular interest in that it allows us to dynamically and intentionally
manipulate our experience, even when our sensory input remains largely
static~\citep[i.e., constant physical stimuli, retinal image, etc.][]{YiEtal06,
OcraEtal99}.

@@

\textbf{JRM STOPPED HERE}
%note: discuss response bias analyses next (also add to supplement)

% We considered each image we probed in the memory phase with respect to whether
% (and how) the participant had experienced and/or attended to that image during
% the presentation phase (Fig.~\ref{fig:familiarity}). For example, if a given
% face image had been presented as part of the composite image on the right of
% the screen following an \texttt{attend to the face on the right} instruction,
% then that face image would have been the focus of the participant's attention
% during one stimulus presentation. The face image on the left (unattended side)
% during that trial would have been captured by feature-based attention, but not
% by location-based attention; and the place image (unattended category) on the
% right during that trial would have been captured by location-based attention,
% but not by feature-based attention. The place image on the left during that
% trial remained outside of the focus of both feature-based and location-based
% attention. In this way, we categorized each of the images that participants
% experienced during the presentation phases of each experiment in terms of
% whether they fell under the scope of feature-based and/or location-based
% attention. We also categorized novel face and place images that we asked
% participants to judge during the memory phase (i.e., images that the
% participant hadn't seen before) as belonging to the attention-cued or uncued
% category. These novel images were intended to serve as a baseline for
% comparison. For example, we sought to measure how people rated the familiarity
% of new images in general. We also used differences in familiarity ratings
% between novel images of the attention-cued versus uncued categories to evaluate
% potential \textit{prospective} affects of modulating the focus of feature-based
% attention.

% \begin{figure*}[tp]
%   \centering
%   \includegraphics[width=0.75\textwidth]{figs/familiarity_by_attention_level}

%   \caption{\textbf{Familiarity ratings for attended, unattended, and novel
%   stimuli.} The bars display the average familiarity ratings participants gave
%   to images from the same category and location as the attention cue (full
%   attention), the same category (but opposite location) as the attention cue
%   (attended category), the same location (but opposite category) as the
%   attention cue (attended location), the opposite category and location as the
%   attention cue (unattended), or novel images. The left panel displays
%   familiarity ratings from the sustained attention condition and the right
%   panel displays familiarity ratings from the variable attention condition. All
%   error bars denote across-participant bootstrap-estimated 95\% confidence
%   intervals. See Figure~\ratingsByCategory~for analogous results broken down by
%   stimulus category.}

% \label{fig:familiarity}
% \end{figure*}

% Participants in both experiments rated stimuli they attended as more familiar
% than unattended or novel stimuli (Fig.~\ref{fig:familiarity}; Sustained
% Attention Experiment: $t(59) = 13.42, p = 1.39 \times 10^{-19}, d = 1.73$;
% Variable Attention Experiment: $t(52) = 12.40, p = 3.70 \times 10^{-17}, d =
% 1.70$). Participants in both experiments also rated attended scene stimuli as
% more familiar than attended face stimuli (Sustained Attention Experiment:
% $t(59) = 7.69, p = 1.85 \times 10^{-10}, d = 0.99$; Variable Attention
% Experiment: $t(52) = 7.08, p = 3.63 \times 10^{-9}, d = 0.97$). These findings
% indicate that aspects of experience that are captured by the focus of attention
% are easier to recognize later, and that attention may preferentially benefit
% memory for some aspects of experience to a greater extent than other aspects of
% experience.

% Participants in the Sustained Attention Experiment experiment also rated
% unattended images that matched the attended category as more familiar than
% unattended images that did not match the attended category (\textit{incidental
% feature-based attention}; $t(59) = 7.17, p = 1.40 \times 10^{-9}, d = 0.93$).
% This pattern did not hold for participants in the Variable Attention Experiment
% ($t(52) = 0.44, p = 0.66, d = 0.06$). Participants in both experiments
% displayed an incidental effect of location-based attention, whereby they rated
% unattended-category images at the attended location as more familiar than
% unattended-category images at the unattended location (Sustained Attention
% Experiment: $t(59) = 4.32, p = 6.05 \times 10^{-5}, d = 0.56$; Variable
% Attention Experiment: $t(52) = 2.57, p = 0.01, d = 0.35$). This indicates that
% aspects of our experience that incidentally overlap with the focus of our
% location-based attention are easier to recognize later, even if those aspects
% do not fall within the intentional focus of our feature-based attention.

% We defined the \textit{memory benefit of feature-based attention} as the
% difference in mean familiarity ratings of attended stimuli and unattended
% stimuli from the attended category versus the mean familiarity ratings of
% unattended stimuli from the unattended category. Similarly, we defined the
% \textit{memory benefit of location-based attention} as the difference in mean
% familiarity ratings of attended stimuli and unattended stimuli from the
% attended location versus the mean familiarity ratings of unattended stimuli
% from the unattended location. Participants in the Sustained Attention
% Experiment displayed a larger memory benefit of feature-based attention than
% location-based attention ($t(59) = 3.25, p = 1.91 \times 10^{-3}, d = 0.42$),
% whereas participants in the Variable Attention Experiment displayed a larger
% memory benefit of location-based attention ($t(52) = 1.70, p = 9.99 \times
% 10^{-2}, d = 0.23$). This appeared to be driven in part by a response bias
% towards attended-category images. For example, relative to the most recent
% attention cue, participants in the Sustained Attention Experiment rated
% attended-category novel images as more familiar than unattended-category novel
% images ($t(59) = 7.04, p = 2.35 \times 10^{-9}, d = 0.91$), whereas
% participants in the Variable Attention Experiment responded similarly to
% attended-category and unattended-category novel images ($t(52) = 0.23, p =
% 0.82, d = 0.03$). The size of this attended-category response bias for novel
% images was reliably smaller than the memory benefits of feature-based attention
% in both experiments (Sustained Attention Experiment: $t(59) = 7.59, p = 2.68
% \times 10^{-10}, d = 0.98$; Variable Attention Experiment: $t(52) = 6.74, p =
% 1.31 \times 10^{-8}, d = 0.93$). This indicates that tuning the focus of
% feature-based attention with respect to how ongoing experience is encoded in
% memory takes longer than the duration of a single stimulus presentation
% (including fixation, roughly 5.5~s). Further, once feature-based attention
% comes ``online,'' it affects how \textit{new} stimuli are processed.

% We also compared the relative sizes of the memory benefits of feature-based
% attention and location-based attention across the two experiments. Participants
% in the Sustained Attention Experiment displayed a larger difference between
% these two effects than did participants in the Variable Attention Experiment
% ($t(111) = 3.48, p = 7.24 \times 10^{-4}, d = 0.66$). Further, the difference
% in mean familiarity ratings of attended-location images versus unattended
% images was greater for participants in the Variable Attention Experiment
% ($t(111) = 2.32, p = 0.02, d = 0.44$). The memory benefit of location-based
% attention reliably distinguished between attended-category versus unattended
% category images for participants in both experiments (Sustained Attention
% Experiment: $t(59) = 9.50, p = 1.68 \times 10^{-13}, d = 1.23$; Variable
% Attention Experiment: $t(52) = 9.72, p = 2.76 \times 10^{-13}, d = 1.33$). This
% indicates that tuning the focus of location-based attention with respect to how
% ongoing experience is encoded into memory occurs faster than the duration of a
% single stimulus presentation.

% We also compared reaction times to probes from the attended versus unattended
% side. Participants in both experiments responded slightly faster on average to
% attended-side probes, although the differences did not reach significance
% (Sustained Attention Experiment: $t(59) = 1.03, p = 0.31, d = 0.13$; Variable
% Attention Experiment: $t(52) = 0.65, p = 0.52, d = 0.09$). Participants in the
% Sustained Attention Experiment displayed a slightly larger (and not
% statistically significant) difference in attended-side versus unattended-side
% reaction times than participants in the Variable Attention Experiment ($t(111)
% = .87, p = 0.38, d = 0.20$). Because none of these reaction time effects were
% statistically reliable, we are unable to draw meaningful conclusions about
% reaction times to the attention probes on the attended versus unattended side,
% nor are we able to draw meaningful conclusions about differences in those
% reaction times that might be due to sustained versus variable attention.

% \begin{figure}[tp]
%   \centering
%   \includegraphics[width=\textwidth]{figures/exps1_2_xgaze.pdf}

%   \caption{\textbf{Horizontal coordinates of visual fixations.} Each violin
%   plot displays a distribution of participant-wise average horizontal gaze
%   positions relative to a presented image. The coordinates have been normalized
%   such that a value of 1.0 denotes the furthest onscreen coordinate from the
%   central fixation point \textit{towards} the direction of the given image, and
%   -1.0 denotes the furthest onscreen coordinate from the central fixation point
%   \textit{away from} the direction of the given image. The gray bars in each
%   panel mark the boundaries of the presented images. The violin plots are
%   broken down by participants' familiarity ratings during the memory phases of
%   each experiment, and by each image's relation to the attention cue while that
%   image appeared onscreen. Panel \textbf{a.} displays fixation data for the
%   Sustained Attention Experiment and Panel \textbf{b.} displays fixation data
%   for the Variable Attention Experiment. Figure~\fixation~displays these
%   results broken down by participant cohort.}

% \label{fig:fixation}
% \end{figure}

% Given that participants in both experiments rated attended-location images from
% both the attended and unattended category as more familiar than
% unattended-location images, we wondered whether this was truly a consequence of
% covert attention, or whether it might be driven solely by where participants
% were looking. In other words, we sought to distinguish the memory effects of
% the focus of covert attention from the focus of visual gaze. We used
% eyetracking to measure the horizontal positions of the participants' visual
% fixations while each pair of composite stimuli appeared onscreen
% (Fig.~\ref{fig:fixation}). For 94\% of presentation trials (Sustained Attention
% Task: $\frac{4531}{4800}$ trials; Variable Attention Task: $\frac{3972}{4240}$
% trials), we collected viable gaze data for analysis. In the remaining trials,
% participant movement or other technical issues corrupted the gaze data. We
% found that, though they generally followed our instruction to keep their gaze
% fixed on the center of the screen, participants in both experiments exhibited a
% slight tendency to look towards the side of the screen they were attending to
% (Sustained Attention Task: $t(59) = 3.61, p = 6.31 \times 10^{-4}$; Variable
% Attention Task: $t(52) = 2.86, p = 6.06 \times 10^{-3}$). When we limited our
% analyses to trials with viable gaze data where participants successfully
% maintained fixation on the center of the screen within 4.28$^\circ$ visual
% angle (i.e., between the rightmost boundary of the left image and the leftmost
% boundary of the right image) we observed (numerical) memory benefits of
% feature-based attention in both experiments (Sustained Attention Experiment:
% $t(59) = 4.73, p = 1.96 \times 10^{-5}, d = 0.67$; Variable Attention
% Experiment: $t(52) = 0.64, p = 0.52, d = 0.09$) and location-based attention in
% the Sustained Attention Experiment, but not the Variable Attention Experiment
% (Sustained Attention Experiment: $t(59) = 0.67, p = 0.51, d = 0.09$; Variable
% Attention Experiment: $t(52) = -0.79, p = 0.43, d = 0.11$). These analyses
% indicate that although the focus of covert attention can affect where people
% are looking, covert attention affects memory encoding beyond what may be
% accounted for by gaze position alone.

% The preceding results indicate that volitional covert attention affects memory,
% and that feature-based and location-based covert attention may come online at
% different timescales. We next sought to examine how long the memory effects of
% feature-based and location-based covert attention persist. To gain additional
% insights into the timecourse of the impact of feature-based and location-based
% attention on memory, we used a sliding window analysis to measure how
% participants' familiarity ratings of attended, partially attended, unattended,
% and novel stimuli varied over the duration of the memory phases of both
% experiments. During each block within the memory phases of each experiment,
% participants made familiarity judgements about a total of 40 images (20 old
% images and 20 novel images). We computed average familiarity ratings for images
% appearing in each position of the memory stimulus sequence (positions 1--40),
% as a function of how (or whether) they were attended during the preceding
% presentation phase (old images) or whether they matched the category of the
% most recent attention cue (novel images), for overlapping 20-image sliding
% windows (Figs.~\ref{fig:familiarity_timecourse_category},
% \ref{fig:familiarity_timecourse_location}, and
% \ref{fig:familiarity_timecourse_novel}). We first calculated these values for
% each participant, and then we averaged across participants.

% \begin{figure*}[tp]
%   \centering
%   \includegraphics[width=1\textwidth]{figures/Cat_Cued_TimeSeries.pdf}

%   \caption{\textbf{Familiarity ratings over time for images that matched the
%   cued or uncued image category.} Each curve reflects the average familiarity
%   ratings for attended, unattended, and novel images (denoted in the legends on
%   the right) within a succession of overlapping 20-image sliding windows. Error
%   ribbons denote 95\% confidence intervals, computed across participants.
%   Panels \textbf{a.} and \textbf{c.} display results from the Sustained
%   Attention Experiment and Panels \textbf{b.} and \textbf{d.} display results
%   from the Variable Attention Experiment. The paired horizontal lines at the
%   bottom of each panel denote timepoints when the given pair of curves was
%   statistically distinguishable (i.e., the topmost line color was statistically
%   greater than the bottommost line color at $\alpha = 0.05$, via a paired
%   two-tailed $t$-test.) The gray lines in Panels b and d reflect familiarity
%   ratings of novel stimuli from both the most recently attended and unattended
%   image categories. Figures~\timecourseCuedCategory~and
%   \timecourseUncuedCategory~display these results broken down by participant
%   cohort.}

% \label{fig:familiarity_timecourse_category}
% \end{figure*}

% Throughout the memory phase, participants from both experiments rated attended
% images as more familiar than category-matched unattended and novel images
% (Fig.~\ref{fig:familiarity_timecourse_category}, top row). This indicates that
% the memory benefits of attending to specific stimulus features (versus
% category-general features) persist for at least as long as the duration of the
% memory phase (2~min). Early in the memory phase, participants in the Variable
% Attention Experiment also rated category-matched unattended images as more
% familiar than novel images. This pattern did not hold for participants in the
% Sustained Attention Experiment. The short-lived boost in familiarity for
% attended-category images at the unattended location following variable (but not
% sustained) attention suggests that the timecourse of feature-based attention's
% impact on memory depends on how long the focus of feature-based attention has
% been held.

% Participants in both experiments also rated unattended-category images at the
% attended location as more familiar than unattended-category images at the
% unattended location and novel unattended-category images
% (Fig.~\ref{fig:familiarity_timecourse_category}, bottom row). This pattern
% persisted throughout the memory phases of both experiments. This supports the
% notion that location-based attention enhances memory in a (partially)
% feature-independent way. Participants in the Variable Attention Experiment also
% rated unattended-category images at the unattended location as (numerically)
% more familiar than novel images (though this pattern was only statistically
% reliable towards the middle of the memory phase). This pattern did not hold for
% participants in the Sustained Attention Experiment. Participants in the
% Sustained Attention Experiment rated unattended-category images (regardless of
% location or novelty status) as less familiar than participants in the Variable
% Attention Experiment (e.g., compare heights of the lines in the bottom left
% versus bottom right panels of Fig.~\ref{fig:familiarity_timecourse_category}).
% This suggests that, after sustained feature-based attention to a particular
% image category, the encodings of stimuli from the unattended category are
% suppressed. This suppression effect appears to persist at least as long as the
% memory phase of the experiment (2~min). Further, the magnitude of this
% suppression effect appears to become stronger following longer-term sustained
% (versus shorter-term variable) feature-based attention.

% \begin{figure*}[tp]
%   \centering
%   \includegraphics[width=\textwidth]{figures/Side_Cued_TimeSeries_2.pdf}

%   \caption{\textbf{Familiarity ratings over time for images that matched the
%   cued or uncued image location.} This figure follows the same format and
%   conventions as Figure~\ref{fig:familiarity_timecourse_category}.
%   Figures~\timecourseCuedLocation~and \timecourseUncuedLocation~display these
%   results broken down by participant cohort.}

% \label{fig:familiarity_timecourse_location}
% \end{figure*}

% We next examined how participants' familiarity ratings varied over the duration
% of the memory phases of each experiment as a function of the attended location.
% Participants in the Sustained Attention Experiment rated attended-category
% images as more familiar, at both the attended and unattended locations
% (Fig.~\ref{fig:familiarity_timecourse_location}, left panels). In contrast,
% participants in the Variable Attention Experiment rated attended-category
% images as more familiar than unattended-category images only at the attended
% locations (Fig.~\ref{fig:familiarity_timecourse_location}, right panels). The
% primary difference in familiarity ratings between Sustained Attention
% Experiment versus Variable Attention Experiment participants was in how they
% rated unattended-category images. Participants in the Variable Attention Task
% rated unattended-category images as more familiar than did participants in the
% Sustained Attention Task; this pattern persisted throughout the memory phase
% and held for images at both the attended and unattended locations (compare
% heights of blue lines in the left versus right panels of
% Fig.~\ref{fig:familiarity_timecourse_location}). This suggests that lower
% familiarity ratings of unattended-category stimuli in the Sustained Attention
% Experiment are due to suppression of the encoding of unattended-category
% features (e.g., as opposed to enhancement of the encoding of attended-category
% features). This suppression effect appears to build over an interval that is
% longer than the duration of a single stimulus presentation (5.5~s).

% \begin{figure*}[tp]
%   \centering
%   \includegraphics[width=1\textwidth]{figures/Novel_Timecourse_Panels.pdf}

%   \caption{\textbf{Familiarity ratings over time for attended-category and
%   unattended-category novel images.} The top panels of this figure follow the
%   same formats and conventions as
%   Figures~\ref{fig:familiarity_timecourse_category} and
%   \ref{fig:familiarity_timecourse_location}. The bottom panels display the
%   differences between the timecourses of familiarity ratings of
%   attended-category and unattended-category novel stimuli. The error ribbons
%   denote 95\% confidence intervals, computed across participants. The dotted
%   horizontal line in each panel denotes a difference of 0-- i.e., the value
%   above which attended-category novel images were rated as more familiar than
%   unattended-category novel images. The black lines in the bottom panels
%   display linear regression fits to the data (error ribbons denote 95\%
%   confidence intervals). Figures~\timecourseNovel~and \timecourseDiff~display
%   these results broken down by participant cohort.}

% \label{fig:familiarity_timecourse_novel}
% \end{figure*}

% Finally, we examined how participants in both experiments rated the familiarity
% of novel images. These ratings provide insights into how modulating the focus
% of attention to \textit{prior} stimuli can affect the perception of
% \textit{future} stimuli. We considered how participants rated novel stimuli
% that matched versus did not match the category (face or scene) of the most
% recent attention cue. Participants in the Sustained Attention Experiment rated
% attended-category novel stimuli as more familiar than unattended-category novel
% stimuli throughout the entire memory phase
% (Fig.~\ref{fig:familiarity_timecourse_novel}a). By contrast, participants in
% the Variable Attention Experiment rated attended-category and
% unattended-category novel images similarly throughout the memory phase
% (Fig.~\ref{fig:familiarity_timecourse_novel}b). The difference in familiarity
% ratings between the two experiments primarily reflects that participants in the
% Variable Attention Experiment rated unattended-category novel images as more
% familiar than did participants in the Sustained Attention Experiment (e.g.,
% compare blue lines in Fig.~\ref{fig:familiarity_timecourse_novel}a versus
% \ref{fig:familiarity_timecourse_novel}b). This may indicate that the processing
% of unattended-category novel stimuli is suppressed following sustained
% attention (e.g., as opposed to enhancing the processing of attended-category
% novel stimuli). The differences in familiarity between attended-category versus
% unattended-category novel images increased slightly but reliably throughout the
% memory phases of both experiments (Fig.~\ref{fig:familiarity_timecourse_novel},
% bottom panels; slope of regression line fit to differences in Sustained
% Attention Experiment: $\beta = 2.05 \times 10^{-3}, p = 1.07 \times 10^{-14} $;
% slope of regression for Variable Attention Experiment: $\beta = 1.54 \times
% 10^{-3}, p = 5.63 \times 10^{-13} $). This might reflect a process whereby the
% suppression of unattended features continues to build for a short duration even
% after the participant stops volitionally focusing on the attended features.

@@

@@

The notion that location-based attention operates at a faster timescale than
feature-based attention is supported by prior work on the deployment of visual
attention~\citep{SotoBlan04, StopEtal07}. Our findings that location-based
attention enhances the processing of attended stimuli whereas feature-based
attention suppresses the processing of unattended stimuli is also consistent
with prior work on location-based attention~\citep[e.g.,][]{IttiKoch01} and
feature-based attention~\citep[e.g.,][]{MoheEtal14}. Our finding that people
better remember attended stimuli also follows prior work on interactions
between attention and memory~\citep{PallWagn02, ChunTurk07, AlyTurk16,
AlyTurk17, WittEtal18, MorrEtal14, BaleEtal19}. Whereas much of this prior work
focused on elucidating the neural basis of these interactions, our work extends
these prior studies by elucidating the specific and separable behavioral
impacts of feature-based attention (inhibition with a slow onset) and
location-based attention (enhancement with a fast onset) on subsequent memory.
Both of these effects persisted throughout the 2~min memory phases of both
experiments. Therefore future work is needed to elucidate the longevity of
these effects beyond 2 minutes.

Another important area for future study concerns how the flow of information
between different brain structures is modulated according to the focus of
volitional attention-- particularly with respect to pathways from primary
sensory regions (e.g., V1, A1) to regions implicated in encoding ongoing
experiences into memory (e.g, medial temporal lobe structures such as the
hippocampus and entorhinal cortex, prefrontal cortex, etc.). For example,
several studies suggest that attention serves to modulate the \textit{gain} of
specific neural circuits~\citep{TreuTruj99, ChanEtal02, EldaEtal13,
SaliThie00}, effectively facilitating or inhibiting the flow of specific neural
representations~\citep{VartEtal07, LaRoEtal14}. Prior work suggests that
feature-based attention may be supported by changes in connectivity with the
thalamus~\citep{Schn11}, whereas location-based attention may be supported by
changes in connectivity with primary visual cortex~\citep{NoudEtal10}. That
feature-based and location-based attention are mediated by different brain
structures may explain why these different aspects of attention operate on
different timescales and affect memory differently. A strong test of this
hypothesis would entail directly measuring neural activity patterns as people
modulate their focus of attention (e.g., using functional magnetic resonance
imaging or electroencephalography), and then using neural decoding
approaches~\citep[e.g.,][]{HaxbEtal01, NormEtal06, MannEtal18} to follow how
neural representations of attended (or unattended) stimuli are transfered from
primary sensory regions, to higher order sensory regions, to memory areas. If
the effects of attention on memory are mediated by changes in network dynamics,
the transmission rates of the representations of attended stimuli from primary
sensory regions to memory areas should be facilitated relative to the
transmission rates of unattended stimuli. Further, variability in these neural
changes (e.g., as a participant focuses their attention with more or less
success) should track with behavioral measures of memorability.

@@

\bibliographystyle{apacite}
\bibliography{./CDL-bibliography/memlab}

>>>>

\newcommand{\gazeLocations}{S1}
\newcommand{\excludedTrials}{S2}
\newcommand{\ratingsByCategory}{S3}
\newcommand{\sequenceLength}{S4}
\newcommand{\responseBias}{S5}

@@

Our brains' cognitive systems detect and exploit patterns in our prior and
ongoing experiences, enabling us to function and adapt in an ever-changing
world. However we do not attend to or treat all types of remembered or incoming
information equally, and our ability to flexibly adapt our thinking and
behaviors can vary markedly with the specific set of concepts or tasks relevant
to a given setting or situation~\citep{ChunTurk07, AlyTurk17, HardNade09,
RangRitc12}. There is also substantial variability across people with respect
to which aspects of experience (sensory, social, emotional, etc.) are noticed,
discriminated between, and acted upon~\citep{HuntEtal89}. This implies that the
same physical (objective) experience may give rise to very different perceived
(subjective) experiences across people~\citep{FreeSimo11, ChanEtal21a}.

The aspects of our experience we attend may be under our volitional control or
may be unconscious or automatic~\citep{JacoEtal92}. Both volitional and
unconscious attention may be expressed overtly, for example through intentional
eye movements~\citep{HoffSubr95} or covertly, without any volitional physical
change~\citep{EngbKlie03}. Prior work has explored the similarities and
differences in the neural basis of overt versus covert
attention~\citep{PosnEtal87, HuntKing03} as well as the behavioral and neural
underpinnings of volitional versus unconscious attention~\citep{DijkAart10} and
their differential effects on memory. There is a general consensus that
sustained volitional attention enhances memory relative to unconscious
attentional processes~\citep{UncaEtal11, TurkEtal13}. However, volitional
attention takes many forms, such as attention to particular spatial locations
or attention to particular visual features or other stimulus properties. How
different \textit{types} of volitional attention combine (or compete) to
enhance memory remains an open question. Volitional covert attention is of
particular interest in that it allows us to dynamically and intentionally
manipulate our experience, even when our sensory input remains largely
static~\citep[i.e., constant physical stimuli, retinal image, etc.][]{YiEtal06,
OcraEtal99}.

@@

Our finding that location-attended items appear to receive a familiarity boost
in the variable attention condition but not the sustained attention condition
is consistent with two possible interpretations. One possibility is that
focusing attention requires just a brief trigger (in this case, an attention
cue), but different forms of attention (in this case, category-based attention
versus location-based attention) require different amounts of time to ``ramp
up'' to full efficacy such that they begin to affect memory encoding. For
example, if category-based attention ramps up more slowly than location-based
attention, this might explain why the relative ordering of category-matched
versus location-matched unattended images changes between the sustained
attention and variable attention conditions (orange and yellow bars in
Fig.~\ref{fig:familiarity}). A second possibility is that each attention cue
provides a ``boost'' to memory encoding for the relevant aspects of one's
experience (e.g., image categories, spatial locations), but the size of the
boost varies across different forms of attention. If so, the \textit{number} of
successive attention cues one receives should predict how effectively the
attended and partially attended images are encoded. We developed a sequence
length analysis to distinguish between these possibilities. For each probed
(target) image that had been presented as a composite image pair, we computed
the number of matching attention cues the participant had received by the time
the given images were presented (up to and including the image's composite
pair). We computed these sequence lengths by defining ``matching'' cues in
three ways: (a) a match means the cues are for the same category \textit{and}
location, (b) a match means that the cues are for the same category, and (c) a
match means that the cues are for the same location. This yielded, for each
target image, an associated count of how many matching attention cues the
participant had received up to and including that image's presentation. As
shown in Figure~\sequenceLength, we used linear regressions of familiarity
rating on sequence lengths to identify potential sequence length effects. For
both experimental conditions, all attention levels, and for all three
approaches to defining matching cue sequence lengths, we found virtually no
reliable associations between cue sequence length and familiarity. This finding
is most parsimonious with the first possiblity mentioned above-- i.e., that
attention may be guided by a brief trigger, but that different forms of
attention may take differently long to affect memory encoding.

Finally, we wondered whether participants' familiarity ratings might be
influenced in part by response bias effects. For example, a participant who had
recently received an ``attend to face images'' cue might rate a probed face
images as more ``familiar'' even if they had no specific memories for the given
image. Indeed, participants in the sustained attention condition do show some
response biases. For example, participants tended to rate novel images as more
familiar if they came from the just-cued category (Fig.~\responseBias A; $t(29)
= 4.371, p < 0.001$). Responses biases are more difficult to evaluate in the
variable attention condition. For example, should cue recency be defined as the
number studied composite image pairs that came between the image whose
familiarity the participant is judging and the most recent save-category cue
(i.e., distance to the nearest same-category probe)? Or might response biases
instead arise when a given category is cued more often near the end of the
just-studied list? We assigned each probed image a label based on these two
measures of category-matched cue recency. We then used linear regressions of
familiarity on these recency measures to identify potential response biases
(Fig.~\responseBias B). Altogether we found no evidence that participants in
the variable attention condition tended to rate images as more familiar solely
due to how recently they had received a same-category attention cue.

@@

\textbf{JRM NOTE: STOPPED HERE}

@@

The notion that location-based attention operates at a faster timescale than
feature-based attention is supported by prior work on the deployment of visual
attention~\citep{SotoBlan04, StopEtal07}. Our findings that location-based
attention enhances the processing of attended stimuli whereas feature-based
attention suppresses the processing of unattended stimuli is also consistent
with prior work on location-based attention~\citep[e.g.,][]{IttiKoch01} and
feature-based attention~\citep[e.g.,][]{MoheEtal14}. Our finding that people
better remember attended stimuli also follows prior work on interactions
between attention and memory~\citep{PallWagn02, ChunTurk07, AlyTurk16,
AlyTurk17, WittEtal18, MorrEtal14, BaleEtal21}. Whereas much of this prior work
focused on elucidating the neural basis of these interactions, our work extends
these prior studies by elucidating the specific and separable behavioral
impacts of feature-based attention (inhibition with a slow onset) and
location-based attention (enhancement with a fast onset) on subsequent memory.
Both of these effects persisted throughout the 2~min memory phases of both
experiments. Therefore future work is needed to elucidate the longevity of
these effects beyond 2 minutes.

Another important area for future study concerns how the flow of information
between different brain structures is modulated according to the focus of
volitional attention-- particularly with respect to pathways from primary
sensory regions (e.g., V1, A1) to regions implicated in encoding ongoing
experiences into memory (e.g, medial temporal lobe structures such as the
hippocampus and entorhinal cortex, prefrontal cortex, etc.). For example,
several studies suggest that attention serves to modulate the \textit{gain} of
specific neural circuits~\citep{TreuTruj99, ChanEtal02, EldaEtal13,
SaliThie00}, effectively facilitating or inhibiting the flow of specific neural
representations~\citep{VartEtal07, LaRoEtal14}. Prior work suggests that
feature-based attention may be supported by changes in connectivity with the
thalamus~\citep{Schn11}, whereas location-based attention may be supported by
changes in connectivity with primary visual cortex~\citep{NoudEtal10}. That
feature-based and location-based attention are mediated by different brain
structures may explain why these different aspects of attention operate on
different timescales and affect memory differently. A strong test of this
hypothesis would entail directly measuring neural activity patterns as people
modulate their focus of attention (e.g., using functional magnetic resonance
imaging or electroencephalography), and then using neural decoding
approaches~\citep[e.g.,][]{HaxbEtal01, NormEtal06b, MannEtal18} to follow how
neural representations of attended (or unattended) stimuli are transfered from
primary sensory regions, to higher order sensory regions, to memory areas. If
the effects of attention on memory are mediated by changes in network dynamics,
the transmission rates of the representations of attended stimuli from primary
sensory regions to memory areas should be facilitated relative to the
transmission rates of unattended stimuli. Further, variability in these neural
changes (e.g., as a participant focuses their attention with more or less
success) should track with behavioral measures of memorability.

@@

\bibliographystyle{apacite}
\bibliography{CDL-bibliography/cdl}

### 2023-03-26 paper/supplemental_materials.tex
finished draft of results section and supplement

%\begin{titlepage}
  \maketitle
%\end{titlepage}

@@

  \label{fig:intersections}
  \end{figure*}

@@

>>>>

\maketitle

@@

@@

\begin{figure*}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/sequence_length_effects}

\caption{\textbf{Effects of attention cue sequence length on familiarity.} 
\textbf{A. Sequence effects (sustained attention condition).} Each heatmap
displays the proportions of images that participants rated at each familiarity
rating (rows), broken down by the number of successive matching attention cues
the participant received at the time the given image was presented as part of a
composite pair (up to and including the image's composite pair; columns). The
trend lines (with bootstrap-estimated 95\% confidence intervals) show
regressions fit to the corresponding heatmap's data ($x$: sequence length; $y$:
familiarity rating). Heatmaps are organized into rows based on how the sequence
lengths were computed (top row: successive attention cues had the same category
and location; middle row: successive attention cues had the same category;
bottom row: successive attention cues had the same location). Columns (and
colors) denote attention levels (as in Figs.~\familiarity~and
\ref{fig:ratings-by-category}). \textbf{B. Sequence effects (variable attention
condition).} This Panel is in the same format as Panel A, but displays results
from the variable attention condition.}

\label{fig:sequence-effects}
\end{figure*}

\begin{figure*}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/response_bias.pdf}

\caption{\textbf{Response bias analyses.} \textbf{A. Sustained attention
condition.} The bars display the mean familiarity ratings participants gave to
novel memory probes (lures) that matched (``True'') versus did not match
(``False'') the prior run's attention-cued category. Error bars denote
across-participant bootstrap-estimated 95\% confidence intervals. \textbf{B.
Variable attention condition.} Each heatmap displays the proportions of images
that participants rated at each familiarity rating (rows), broken down by
metrics related to recent same-category attention cues participants received on
the prior run (columns). The trend lines (with bootstrap-estimated 95\%
confidence intervals) show regressions fit to the corresponding heatmap's data
($x$: metrics related to same-category cues; $y$: familiarity rating). The top
panels's columns reflect the temporal distances to the most recent
same-category attention cues, measured in numbers of presented items (where 1
denotes the final item on the most recent run). The bottom panel's columns
denote the recency-weighted average numbers of same-category cues presented on
the most recent run. Each column of matrices corresponds to a different
attention level, as in Figures~\familiarity~and \ref{fig:ratings-by-category}.
\textbf{C. Weights by recency.} The panel displays the weights used to compute
recency-weighted numbers of same-category cues on the most recent run. The
$x$-axis displays the presentation positions of each attention cue, and the
$y$-axis displays the weight assigned to same-category cues at the given
presentation position. The weights are first computed using $w = \max\left[1 -
\exp\{-\frac{x}{\tau}\} , \epsilon \right]$, where $\tau = 2$, $\epsilon =
0.05$, and $x$ is the cue's presentation position. To obtain the $y$-values
shown in the Panel (and used in Panel B), the weights for $x \in \left[0, 1, 2,
..., 9\right]$ are normalized to sum to 1.}

\label{fig:response-bias}
\end{figure*}
